Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2023-11-29 article.qmd
Extend figure captions

Symbolite is a lightweight symbolics package to create algebraic mathematical expressions.
They can be inspected and compiled to different backends.
Currently, we have implementations for
NumPy [@harrisArrayProgrammingNumPy2020];
Numba [@lamNumbaLLVMbasedPython2015], a Just-in-Time (JIT) compiler to LLVM;
SymPy [@meurerSymPySymbolicComputing2017], a library for symbolic mathematics;
and JAX [@jax2018github], a library that support automatic differentiation
and compilation to GPUs and TPUs;
but it is easy to add new backends.

@@

First-order system of an exponential decay.
:::

@@

Second-order system of an harmonic oscillator,
where $\dot{x}$ is named $v$.
:::

::: {#fig-composition layout-ncol=3}
```{python}
class BigModel(System):
    x: Variable = initial(default=1)
    subsystem = Decay(x=x)
    independent = Decay(x=2)

@@

First-order system of two exponential decays,
where $y$ is named $\text{independent.x}$.
:::

@@

::: {#fig-reaction layout-ncol=2}
```{python}
class Model(Compartment):
    """2A -> B"""

@@

$$
\begin{align}
&\begin{cases}
    \frac{dA}{dt} = -2 \\
    \frac{dB}{dt} = +1
\end{cases}
\\\\
&\begin{cases}
    A(0) &= 1 \\
    B(0) &= 0
\end{cases}
\end{align}
$$

Reaction system $2A \rightarrow B$ with rate $1$.
:::

::: {#fig-mass-action layout-ncol=2}
```{python}
class Model(Compartment):
    """2A -> B"""

@@

$$
\begin{align}
&\begin{cases}
    \frac{dA}{dt} = -2 A^2 \\
    \frac{dB}{dt} = +1 A^2
\end{cases}
\\\\
&\begin{cases}
    A(0) &= 1 \\
    B(0) &= 0
\end{cases}
\end{align}
$$

Mass-Action system $2A \rightarrow B$ with rate $1 A^2$.
:::

Examples of similar reaction systems transforming $2A \rightarrow B$.
For mass-action systems, the rate is proportional to the reactants.
::::

For reaction networks,
we are interested in first-order differential equations
relating the rate of change of the species.
To facilitate defining these models,
SimBio defines `Species`,
which comprises of a `poincare.Variable` and an stoichiometry,
and `Reaction`,
which transform reactants into products taking into account the stoichiometry (@fig-reaction).
It also includes `MassAction`, a `Reaction` subclass,
which automatically adds the reactants as part of the rate law (@fig-mass-action).
Several commonly used reactions are predefined as `MassAction` subclasses.
Additionally, SimBio supports importing from and exporting to SBML,
or directly from BioModels (@fig-simbio-io).

>>>>

Symbolite is a lightweight symbolics package to create algebraic mathematical expressions.
They can be inspected and compiled to different backends.
Currently, we have implementations for
NumPy [@harrisArrayProgrammingNumPy2020];
Numba [@lamNumbaLLVMbasedPython2015], a Just-in-Time (JIT) compiler to LLVM;
SymPy [@meurerSymPySymbolicComputing2017], a library for symbolic mathematics;
and JAX [@jax2018github], a library that support automatic differentiation
and compilation to GPUs and TPUs;
and it is easy to add new backends.

@@

First-order system of an exponential decay.
The variable `x` stores the initial condition for $x$,
and the variable `eq` stores the rate equation for $x$.
:::

@@

Second-order system of an harmonic oscillator,
where the variable `v` stores the derivative $\dot{x}$
and the rate equation is specified for the derivative `v`.
:::

::: {#fig-composition layout-ncol=3}
```{python}
class BigModel(System):
    x: Variable = initial(default=1)
    linked = Decay(x=x)
    independent = Decay(x=2)

@@

First-order system of two exponential decays by composition of the `Decay` system.
The subsystem `linked` has a reference to the outer variable `x`,
while the subsystem `independent` defines a new variable with initial condition §2$,
which on the corresponding mathematical expression was named $y$.
:::

@@

::: {layout-ncol=2}
```{python}
class Model(Compartment):
    """2A -> B"""

@@

$$
\begin{align}
&\begin{cases}
    \frac{dA}{dt} = -2 \\
    \frac{dB}{dt} = +1
\end{cases}
\\\\
&\begin{cases}
    A(0) &= 1 \\
    B(0) &= 0
\end{cases}
\end{align}
$$
:::

::: {layout-ncol=2}
```{python}
class Model(Compartment):
    """2A -> B"""

@@

$$
\begin{align}
&\begin{cases}
    \frac{dA}{dt} = -2 A^2 \\
    \frac{dB}{dt} = +1 A^2
\end{cases}
\\\\
&\begin{cases}
    A(0) &= 1 \\
    B(0) &= 0
\end{cases}
\end{align}
$$
:::

A reaction system for species $A$ and $B$ with initial conditions $1$ and $0$, respectively.
A single reaction transforming $2A$ into $B$ is saved in variable `r`.
The rate $1$ is specified directly for `Reaction`,
and is proportional to the reactants for `MassAction`.
::::

For reaction networks,
we are interested in first-order differential equations
relating the rate of change of the species.
To facilitate defining these models,
SimBio defines `Species`,
which comprises of a `poincare.Variable` and an stoichiometry,
and `Reaction`,
which transform reactants into products taking into account the stoichiometry (@fig-simbio).
It also includes `MassAction`, a `Reaction` subclass,
which automatically adds the reactants as part of the rate law (@fig-simbio).
Several commonly used reactions are predefined as `MassAction` subclasses.
Additionally, SimBio supports importing from and exporting to SBML,
or directly from BioModels (@fig-simbio-io).
